---
title: "Polishing plot output: labels, watermarks, and session options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polishing plot output: labels, watermarks, and session options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align='center'
)
```

```{r setup, echo=FALSE, include = FALSE}
library(dplyr)
library(xpose)
library(xpose.xtras)
```

## Introduction

A model diagnostic is rarely just a diagnostic once it leaves an R session: it becomes a figure in a report, and figures in reports want consistent titles, an obvious "this is a draft" stamp while a model is still in flux, and a predictable place to land on disk. Doing all of that by hand, plot-by-plot and model-by-model, gets old fast.

`xpose.xtras` adds a small, composable toolkit for this last mile:

- `apply_default_labs()` / `set_default_labs()` fill in (or overwrite) plot titles, subtitles, captions and tags.
- `add_watermark()` / `set_default_watermark()` stamp a plot with draft/preliminary/confidential-style text.
- `ggsave_xp()` saves a plot, applying the above first, with a swappable save backend.

All three read from the same kind of layered defaults, resolved in the same order every time: a session-wide R option, then a per-model (`xpdb`) default, then whatever is passed directly to the function call. `set_xtras_options()` and `get_xtras_option()` round this out by giving the whole `xpose.xtras.*` option family one place to be set and inspected.

None of this is wired into `print()` -- every function here is opt-in, called explicitly on a plot (or a save call) when you want it.

## Default labels

`apply_default_labs()` takes a `ggplot`/`xpose_plot` object and fills in any of `title`/`subtitle`/`caption`/`tag` that aren't already set, using (in increasing precedence) the `xpose.xtras.default_labs` option, an `xpdb`-level default, and anything passed directly.

The simplest use is a session-wide option, handy for something that should apply to every plot in a script or report -- a draft caption, say:

```{r default_labs_option}
options(xpose.xtras.default_labs = list(caption = "DRAFT -- do not distribute"))

p <- dv_vs_ipred(xpdb_ex_pk, quiet = TRUE)
p_labelled <- apply_default_labs(p)
p_labelled
```

Because `dv_vs_ipred()` already sets its own caption, that one wasn't touched -- `apply_default_labs()` only fills in labels that are still empty. Passing `overwrite = TRUE` replaces existing labels too, and the value can use the same `@keyword` placeholders `xpose::parse_title()` understands elsewhere in the package:

```{r default_labs_overwrite}
apply_default_labs(p, caption = "Run @run -- @nobs observations", overwrite = TRUE)
options(xpose.xtras.default_labs = NULL)
```

A single global caption is a blunt instrument once more than one model is in play. `set_default_labs()` attaches label defaults to a specific `xpdb` instead, and those take precedence over the option -- useful when different models in the same report need different framing:

```{r default_labs_xpdb}
xpdb_draft <- xpdb_x %>%
  set_default_labs(caption = "Model @run -- interim, subject to change")

dv_vs_ipred(xpdb_draft, quiet = TRUE) %>%
  apply_default_labs(xpdb = xpdb_draft, overwrite = TRUE)
```

`apply_default_labs()` can't reach back into the `xpdb` a plot was built from on its own -- `xpose`'s plotting functions only carry a reduced summary forward onto the plot object, not the full `xpdb` -- so the `xpdb` argument has to be supplied explicitly whenever the `xpdb`-level tier (or `@keyword` resolution) should be used. Whatever is passed directly to `apply_default_labs()` still wins over both tiers:

```{r default_labs_direct}
dv_vs_ipred(xpdb_draft, quiet = TRUE) %>%
  apply_default_labs(caption = "Final for submission", xpdb = xpdb_draft, overwrite = TRUE)
```

## Watermarks

`add_watermark()` overlays large, semi-transparent, rotated text across a plot -- the kind of stamp that makes it obvious at a glance that a figure isn't final yet.

```{r watermark_basic}
p %>%
  add_watermark()
```

`label`, `colour`, `alpha`, `size`, `angle` and `fontface` are all adjustable:

```{r watermark_custom}
p %>%
  add_watermark(label = "PRELIMINARY", colour = "firebrick", alpha = 0.15, angle = 20)
```

The watermark is added as an ordinary `ggplot2` layer, so it survives faceting and pagination -- one watermark per panel, automatically:

```{r watermark_facets}
dv_vs_ipred(xpdb_ex_pk, quiet = TRUE, facets = "SEX") %>%
  add_watermark(label = "DRAFT")
```

Because the watermark is drawn at a fixed rotation, it can visually merge into a plot that already has a strong diagonal of its own -- like the line of unity on a `dv_vs_pred()` plot. The default 30 degree angle sits close to that line and gets lost among the points around it:

```{r watermark_angle_unity_default}
pred_plot <- dv_vs_pred(xpdb_ex_pk, quiet = TRUE)

pred_plot %>%
  add_watermark(alpha = 0.6)
```

Crossing the other way keeps the watermark clearly legible instead, without competing with the plot's own diagonal -- this is exactly the kind of case `angle` (and `alpha`/`size`/`colour`) exist for:

```{r watermark_angle_unity_adjusted}
pred_plot %>%
  add_watermark(alpha = 0.6, angle = -45)
```

Like the label functions, watermark settings resolve from the `xpose.xtras.default_watermark` option, then an `xpdb`-level default set with `set_default_watermark()`, then whatever is passed directly -- so a whole project can default to (say) a red "CONFIDENTIAL" stamp without repeating those arguments at every call site:

```{r watermark_option}
options(xpose.xtras.default_watermark = list(label = "CONFIDENTIAL", colour = "firebrick"))
p %>%
  add_watermark()
options(xpose.xtras.default_watermark = NULL)
```

## Saving with `ggsave_xp()`

`ggsave_xp()` is a drop-in, `ggplot2::ggsave()`-compatible way to write a plot to disk: same `plot`/`filename`/`path`/`width`/`height` arguments, but it applies `apply_default_labs()` first (`apply_labs = FALSE` to skip that), and `path`/`width`/`height` fall back to the `xpose.xtras.save_dir`/`save_width`/`save_height` options when not supplied.

```{r ggsave_xp_basic}
out_dir <- tempdir()

saved_path <- ggsave_xp(p, filename = "dv_vs_ipred.png", path = out_dir, width = 6, height = 4)
basename(saved_path)
```

That's a real file on disk -- here's what actually got written:

```{r ggsave_xp_basic_preview, echo = FALSE, out.width = "80%"}
knitr::include_graphics(saved_path)
```

`filename`/`path` can also use `@keyword` placeholders, expanded the same way `xpose::xpose_save()` does it:

```{r ggsave_xp_keywords}
run_path <- ggsave_xp(dv_vs_ipred(xpdb_ex_pk, quiet = TRUE), filename = "@run_@plotfun.png", path = out_dir)
basename(run_path)
```

What actually does the saving is itself swappable, via `save_fun` -- the default is `ggplot2::ggsave()`, but anything sharing its signature works, including third-party wrappers like `reportifyr::ggsave_with_metadata()`. As a self-contained illustration (so this vignette doesn't need to depend on a third-party package just to demonstrate the idea), here's a `save_fun` that writes a plain-text note alongside the image:

```{r ggsave_xp_custom_backend}
save_with_note <- function(plot, filename, path = NULL, width, height, note = "", ...) {
  saved <- ggplot2::ggsave(filename = filename, plot = plot, path = path, width = width, height = height, ...)
  writeLines(note, sub("\\.[^.]+$", ".txt", saved))
  saved
}

annotated_path <- ggsave_xp(
  p, filename = "dv_vs_ipred_annotated.png", path = out_dir,
  save_fun = save_with_note, note = "Generated for the Q3 interim analysis."
)

# the .png was written by ggplot2::ggsave() as usual; the .txt is new
readLines(sub("\\.png$", ".txt", annotated_path))
```

Any function accepting `plot`/`filename`/`path`/`width`/`height` (plus whatever else it needs via `...`) can be dropped in this way -- there's nothing `xpose.xtras`-specific about `save_fun` itself.

## Tying it together with session options

Setting several of the options above one at a time is easy to get wrong -- a typo in an option name just silently does nothing, since `options()` doesn't validate names. `set_xtras_options()` is a small, validated wrapper for the whole `xpose.xtras.*` family (prefix added automatically), good for pinning a project's defaults once at the top of a script:

```{r set_xtras_options}
set_xtras_options(
  default_labs = list(caption = "DRAFT"),
  default_watermark = list(label = "DRAFT"),
  save_dir = tempdir(),
  save_width = 6,
  save_height = 4
)

dv_vs_ipred(xpdb_ex_pk, quiet = TRUE) %>%
  apply_default_labs() %>%
  add_watermark()
```

The same family also covers a session-wide default theme -- `gg_theme`/`xp_theme` -- picked up automatically the first time an `xpose_data` object is converted with `as_xpdb_x()`, so a project look can be set once instead of calling `xpose::update_themes()` on every model:

```{r set_xtras_options_theme}
set_xtras_options(gg_theme = theme_bw)

xpdb_ex_pk %>%
  as_xpdb_x() %>%
  dv_vs_ipred(quiet = TRUE)
```

`get_xtras_option()` reports, for a given option and (optionally) a specific `xpdb`, what's currently set at each tier and which one would win:

```{r get_xtras_option}
xpdb_labelled <- xpdb_x %>%
  set_default_labs(caption = "Model-specific caption")

get_xtras_option("default_labs", xpdb_labelled)
```

```{r set_xtras_options_reset, include = FALSE}
set_xtras_options(
  default_labs = NULL, default_watermark = NULL,
  save_dir = NULL, save_width = NULL, save_height = NULL,
  gg_theme = NULL
)
```

The full list of recognized options -- what each one does, and which ones have an `xpdb`-level counterpart -- is documented on `?set_xtras_options`.
