---
title: "Useful uncommon plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Useful uncommon plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align='center'
)
```

```{r setup, echo=FALSE, include = FALSE}
library(dplyr)
library(xpose)
library(xpose.xtras)
```

## Introduction

The `xpose.xtras` package attempts to bring old favorites back to the `xpose` framework from its predecessor `xpose4`, and those are named as such to allow easy access to documentation. This vignette brings focus to the visualizations in this package that are unlikely to be used for most projects, but are still helpful to have when they are needed. The underlying functionality for these plots are also useful to generalize to similar use cases that will continue to have the benefit of being within the `xpose` framework.

## Model-averaged plots

Model-averaging can be useful when two or more models can describe different aspects of the data or the pharmacology, but for various reasons a mixture model or other population approach would be inadequate. It can also be helpful when multiple models have been developed for various populations and a new population being fit does not necessarily consist entirely of any previously fitted population.

This package facilitates the generation of model-averaged diagnostics. The approach is rudimentary and experimental, merely creating an averaged version of an `xpose_data` object from an `xpose_set`. As referenced in the documentation, the Model Selection and Model Averaged Algorithms from Uster et al. are implemented to do this averaging. Because both algorithms require individual objective functions, there is an argument to automatically apply `backfill_iofv()`. 

```{r modavg, fig.width=unit(7,"in"), fig.height=unit(5,"in")}
pheno_set %>%
  ipred_vs_idv_modavg(auto_backfill = TRUE, quiet=TRUE)
```

The default title, subtitle and caption for these experimental figures are rough, especially for large sets. Changes for better appearance should be expected in the future.

## Categorical dependent variables

Categorical DVs are frequently modeled but there are a variety of methods used to visually diagnose these modeled endpoints, typically model-specific. Along with adding generic support for categorical DVs, `xpose.xtras` also adds a few plots to diagnose models developed for them.

To use these diagnostics, a model for a categorical DV needs to have the column for that DV stated (if it is "DV" that needs to be ripped away from the `dv` variable type), and have a column predicting the likelihood or probability of that DV having a certain value. An example using an M3 model is below and in the documentation.

```{r m3, fig.width=unit(7,"in"), fig.height=unit(5,"in")}
pkpd_m3 %>%
  # Need to ensure var types are set
  set_var_types(catdv=BLQ,dvprobs=LIKE) %>%
  # Set probs ("LIKE is the probability tht BLQ is 1")
  set_dv_probs(1, 1~LIKE, .dv_var = BLQ) %>%
  # Optional, but useful to set levels
  set_var_levels(1, BLQ = lvl_bin()) %>%
  catdv_vs_dvprobs(quiet=TRUE)
```

It is expected that there will be a somewhat sigmoidal or at least up- and right-ward relationship between the two sets of observations in an adequate model, but as with all diagnostics the interpretation for this plot is subjective to an extent.

All plots for categorical DVs have at most two sets of data. It is untested what would happen if all observations fell into one category, but it is unlikely to produce a good model. This is relevant because the `catdv` functions can still be used for models that have multiple levels, such as the vismodegib muscle spasm model (from Lu et al.) built into the examples. For these models, the plot is essentially dichotomizing the probability of one observation compared to the probability of not that observation,


```{r vismo, fig.width=unit(7,"in"), fig.height=unit(5,"in")}
vismo_xpdb <- vismo_pomod  %>%
  set_var_types(.problem=1, catdv=DV, dvprobs=matches("^P\\d+$")) %>%
  set_dv_probs(.problem=1, 0~P0,1~P1,ge(2)~P23)
vismo_xpdb %>%
  catdv_vs_dvprobs(quiet=TRUE)
vismo_xpdb %>%
  catdv_vs_dvprobs(cutpoint = 2, quiet=TRUE)
vismo_xpdb %>%
  catdv_vs_dvprobs(cutpoint = 3, quiet=TRUE)
```

None of the examples can be used to demonstrate this easily, but these plots can also be model-averaged using the generic `plotfun_modavg` function.

## Waterfall and objective function trends

The more common needs for an `xpose_set` include model-building tables, covariate testing and visualization. However, there are occasions where a visual description of what can be shown in these tables can be useful.

The default waterfall plot compares scales the change in parameter values, which is intended to make relative comparisons the focus. This is especially beneficial in comparing changes in empirical Bayes estimates (EBEs).

```{r waterfall, fig.width=unit(7,"in"), fig.height=unit(5,"in")}
pheno_set %>%
  eta_waterfall(run3,run6, quiet=TRUE)
```

Waterfalls can also be used as an alternative to shark plots. Scaling in that case is off by default.

```{r waterfall2, fig.width=unit(7,"in"), fig.height=unit(5,"in")}
pheno_set %>%
  focus_qapply(backfill_iofv) %>%
  iofv_waterfall(run3,run6, quiet=TRUE)
```

To track iOFV changes over multiple models, another plot can be used.

```{r ofv_change, fig.width=unit(7,"in"), fig.height=unit(5,"in")}
iofv_vs_mod(pheno_set, auto_backfill = TRUE, quiet=TRUE)
```
